\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

% ---------------------------------------------------------------
% Human acknowledgements (customize as needed)
% ---------------------------------------------------------------

This book exists because of the many colleagues, reviewers, and
practitioners who shared their hard-won operational knowledge. I am
especially grateful to Dr. John Santerre for their close technical review of
early drafts, and to Dr. Julie Penzotti for extended discussions that shaped the
Ishtar~AI case study into something concrete enough to teach from.
Any errors that remain are mine.

I thank the team at Ishtar AI for the environment in which many
of the ideas in this book were first tested under real production
conditions, and the editorial team at Springer Nature for their
patience and guidance through the writing process.

This book is dedicated to the memory of my son, John David Stroud, and his 
brothers, Spencer and Wilson Stroud.

% ---------------------------------------------------------------
% Declaration of AI-assisted content
% Required by Springer Nature Manuscript Guidelines for monographs.
% Customize to reflect your actual use. Do not remove.
% ---------------------------------------------------------------

\section*{Declaration of AI-Assisted Content}

In accordance with Springer Nature's policy on the use of large language
models in authored content, I disclose the following:

Portions of this manuscript were developed with the assistance of large
language model tools, specifically Anthropic Claude
Opus 4.7, used in the following capacities:

\begin{itemize}
  \item Drafting and revision of prose, with all substantive content
        reviewed, verified, and edited by the author.
  \item Generation of illustrative code listings and LaTeX source for
        figures typeset via TikZ and PGFPlots. All figures in this
        book are vector drawings; no figures were produced using
        generative-AI image synthesis.
  \item Structural compliance auditing of the manuscript against
        Springer Nature's monograph formatting requirements.
\end{itemize}

The author retains full responsibility for the accuracy, originality,
and integrity of all content in this book. No AI tool is listed as an
author, and no AI-generated content was incorporated without author
review. Citations, technical claims, and attributions have been
verified by the author against primary sources.
